Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-10-17 manuscript/plos/review_1.tex
changed to past tense

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Biological networks and GWAS: comparing and combining network methods to understand the genetics of familial breast cancer susceptibility in the GENESIS study} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Héctor Climente-González\textsuperscript{1,2,3,4*}, 
Christine Lonjou\textsuperscript{1,2,3}, 
Fabienne Lesueur\textsuperscript{1,2,3},
Dominique Stoppa-Lyonnet\textsuperscript{5,6,7\textcurrency}, 
Nadine Andrieu\textsuperscript{1,2,3}, 
Chloé-Agathe Azencott\textsuperscript{3,1,2},
with the GENESIS study group\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Institut Curie, PSL Research University, F-75005 Paris, France;\\
\textbf{2} INSERM, U900, F-75005 Paris, France;\\
\textbf{3} MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
\textbf{4} RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, Japan; \\
\textbf{5} Service de Génétique, Institut Curie, F-75005 Paris, France;\\
\textbf{6} INSERM, U830, F-75005 Paris, France;\\
\textbf{7} Université Paris Descartes.\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
% Network methods provide a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses.
% They discover susceptibility genes by jointly consider the statistical association between genetic variation and a phenotype, measured in a genome-wide association study (GWAS), and the biological context of each gene, represented as a network.
Network approaches to disease use molecular networks, which model functional relationships between the molecules in a cell, to generate hypotheses about the genetics of complex diseases.
Among them, several approaches jointly consider gene scores, representing the association between each gene and the disease, and the biological context of each gene, modeled by a network.
%These approaches can select low-scoring genes if they connect multiple high-scoring genes.
Here, we obtain gene scores from genome-wide association studies (GWAS) and study six network approaches on 
% Hence, a network method could select a gene with a high P-value of association if it connects multiple low P-value genes in the network. In this work, we studied six network methods which identify subnetworks with high scores of GWAS association with a phenotype. This allows them to give more compelling results than standard SNP- and gene-level GWAS analyses, recovering causal subnetworks tightly related to cancer susceptibility.
% We apply them to
GENESIS, a GWAS on French women with non-BRCA familial breast cancer% and no pathogenic variants in \emph{BRCA1} and \emph{BRCA2}
. We provide a critical comparison of these six methods, discussing the impact of their mathematical formulation and of their parameters. % on % . Additionally, we performed an in-depth benchmarking with respect to
% the size and predictive power of their solutions, as well as on their stability and runtimes.
% Importantly,
Using a molecular network yields more compelling results than standard % SNP- and gene-level
GWAS analyses. Indeed, %recovering causal subnetworks tightly related to cancer susceptibility.
we find significant overlaps between our solutions and the genes identified in the largest GWAS on breast cancer susceptibility. % Since most of the methods produce reasonable solutions, we
We further
propose to combine these solutions into a consensus network, %containing the genes selected by at least two methods. This aggregation
which
brings further insights. The consensus network contains \emph{COPS5}, a gene related to multiple hallmarks of cancer, and 14 of its neighbors. The main drawback of network methods % with respect to conventional $\chi^2$-based GWAS
is that they are not robust to small perturbations in their inputs. %To compensate for this instability, we
We therefore propose a stable consensus solution, formed by the most consistently selected genes in multiple subsamples of the data. In GENESIS, it is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than most genes. The network is organized around \emph{CUL3}, which is involved in the regulation of several genes linked to cancer progression. In conclusion, we showed how network methods help overcome the lack of statistical power of GWAS and improve their interpretation. Project-agnostic implementations of all methods are available at \url{https://github.com/hclimente/gwas-tools}. % to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an appealing explanation: only a limited number of core genes are directly involved in the disease; but gene functions are deeply interrelated so that many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we apply and compare six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expression is correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,pinero_uncovering_2016}: they tend to occupy central positions in the network, interconecting different modules TODO. Therefore, studying the neighborhood of disease-associated genes is effective at identifying new ones that are involved in the same biological functions \cite{huang_systematic_2018}.

@@

In this work, we analyze the application of these six network methods on GWAS data. They use different interpretations of the omnigenic model, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we use these network methods to identify additional breast cancer susceptibility genes. Lastly, we carry out a comparison of the solutions obtained by the different methods, and aggregate them to obtain a consensus solution of predisposition to familial breast cancer.

@@

We conducted association analyses in the GENESIS dataset (Section~\ref{methods:data}) at both SNP and gene levels (Section~\ref{methods:conventional}). Two genomic regions had a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (\nameref{sfig:snp_gene_manhattan}A). The former overlaps with gene \emph{FGFR2}; the latter with \emph{CASC16} the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1} and \emph{BRCA2} carrier studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. In GENESIS only \emph{FGFR2} was significantly associated with breast cancer at the gene-level (\nameref{sfig:snp_gene_manhattan}B).

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (Benjamini-Hochberg [BH] correction adjusted P\=/value $<$ 0.05), we obtained six solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, one for SConES GI (which works at the SNP level), and the consensus.

@@

The topologies of the six solutions differ as well (Fig~\ref{fig:solution_overview}C), as measured by the median centrality  and the number of connected components (Table~\ref{tab:gene_solutions}). Only two methods yield more than one connected component: SConES, as described above, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have fewer than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM}.

Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences between the solutions of the different methods. It linked different parts of SigMod's solution to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~$<$~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) is also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). However, it additionally includes submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~$<$~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) are involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~$<$~0.04).

\begin{table}[!ht]
  \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
  \centering
  \caption{
  {\bf Summary statistics on the solutions of multiple network methods on the PPIN. The first row contains the summary statistics on the whole PPIN.}}
  \begin{tabular}{lrrrrrr}
  {\bf Network } & {\bf \# genes } & {\bf \# edges } & {\bf \# components } & {\bf \mean{Betweenness} } & {\bf \median{P}\textsubscript{gene} } & {\bf \# genes in consensus } \\
  \thickhline
  HINT HT           & 13\,619 & 142\,541  & 15  & 16\,706   & 0.46  & 93/93 \\
  \hline
  dmGWAS            & 194     & 450       & 1   & 49\,115   & 0.19  & 55/93 \\
  heinz             & 4       & 3         & 1   & 113\,633  & 0.001 & 4/93  \\
  HotNet2           & 440     & 374       & 130 & 7\,739    & 0.048 & 63/93 \\
  LEAN              & 0       & 0         & 0   & -         & -     & 0/93  \\
  SConES GI         & 0 (1)   & 0         & 0   & -         & -     & 0/93  \\
  SigMod            & 142     & 249       & 11  & 92\,603   & 0.008 & 84/93 \\
  \hline
  Consensus         & 93      & 186       & 21  & 50\,737   & 0.006 & 93/93 \\
  Stable consensus  & 68      & 49        & 32  & 94\,854   & 0.005 & 43/93 \\
  \end{tabular}
  \begin{flushleft} \textbf{\# genes}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. {\bf \# components }: number of connected components. \textbf{\mean{Betweenness}}: mean betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median VEGAS2 P\=/value of the selected genes. \textbf{\# genes in consensus}: Number of genes in common between the method's solution, and the 93 genes in the consensus solution.
  \end{flushleft}
  \label{tab:gene_solutions}
  \end{adjustwidth}
\end{table}

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant solution (BH adjusted P\=/value~$<$~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)}~Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors (Sections \ref{methods:algorithm_comparison} and \ref{methods:snp2gene}). \textbf{(B)}~VEGAS2 P-values of the genes in the PPIN not selected by any network method (12\,213), and of those selected by 1 (575), 2 (73), or 3 (20) methods. \textbf{(C)}~Solution networks produced by the different methods. \textbf{(D)}~Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes).}
  \label{fig:solution_overview}
\end{figure}

Despite their differences, there are additional common themes. All obtained solutions have lower association P\=/values than the whole PPIN (median VEGAS2 P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}D). This illustrates the trade-off between controlling for type I error and biological relevance. However, there are nuances between solutions in this regard: heinz strongly favors genes with lower P\=/values, while dmGWAS is less conservative (median VEGAS2 P\=/values 0.0012 and 0.19, respectively); SConES tends to select whole LD-blocks; and HotNet2 and SigMod are less likely to select low scoring genes.

Additionally, the solutions presented other desirable properties. First, five of them were enriched in known breast cancer susceptibility genes (consensus, dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value$<$0.03). Second, the genes in four solutions displayed on average a higher betweenness centrality than the rest of the genes, a difference that is significant in four solutions (consensus, dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value$<$1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non-associated SNPs.

@@

Despite their shared properties, the differences between the solutions of the different methods suggest that each of them captures different aspects of cancer susceptibility. Indeed, out of the 668 genes that are selected by at least one method, only 93 are selected by at least two, 20 by three, and none by four or more. Encouragingly, the more methods selected a gene, the higher its association score to the phenotype (Fig \ref{fig:solution_overview}B), a relationship that plateaus at 2. Hence, to leverage on their strengths and compensate their respective weaknesses, we built a consensus solution that captures the genes shared among at least two solutions (Section~\ref{methods:methods}). This solution (Fig~\ref{fig:consensus}) contains 93 genes and exhibits the aforementioned properties of the individual solutions: enrichment in breast cancer susceptibility genes and higher betweenness centrality than the rest of the genes.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_3.pdf}
  \caption{ {\bf Consensus solution on GENESIS (Section~\ref{methods:methods}) }. \textbf{A} ~Manhattan plots of genes; in black, the ones in the consensus solution. The Bonferroni threshold is indicated by a red line (1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{B} Consensus network. Each gene is represented by a pie chart, which shows the methods that selected it. We enlarged the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset (Section~\ref{methods:bcac}). \textbf{C} The nodes are in the same disposition as in panel B, but every gene name is indicated. We colored in pink the names of genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset}
  \label{fig:consensus}
\end{figure}

A pathway enrichment analysis of the genes in the consensus solution also shows similar pathways as the individual solutions (\nameref{stab:consensus_pwy}). We found two involved mechanisms: \emph{mitochondrial translation} and \emph{attenuation phase}. The former is supported by genes like \emph{MRPS30} (VEGAS2 P\=/value~=~0.001), which encode a mitochondrial ribosomal protein and was also linked to breast cancer susceptibility \cite{quigley_5p12_2014}. Interestingly, increased mitochondrial translation has been found in cancer cells \cite{Yu2016Repositioning}, and its inhibition proposed as a therapeutic target. With regards to the attenuation phase of heat shock response, it involves three Hsp70 chaperones: HSPA1A, HSPA1B, and HSPA1L. The genes encoding these proteins are all near each other at 6p21, in the region known as HLA. In fact, out of the 22 SNPs that map to any of these three genes, 9 map to all of them, and 4 to two, making it hard to disentangle their effects. \emph{HSPA1A} was the most strongly associated gene (VEGAS2 P\=/value~=~8.37~\texttimes{}~10\textsuperscript{-4}).

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Biological networks and GWAS: comparing and combining network methods to understand the genetics of familial breast cancer susceptibility in the GENESIS study} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Héctor Climente-González\textsuperscript{1,2,3,4*}, 
Christine Lonjou\textsuperscript{1,2,3}, 
Fabienne Lesueur\textsuperscript{1,2,3},
Dominique Stoppa-Lyonnet\textsuperscript{5,6,7\textcurrency}, 
Nadine Andrieu\textsuperscript{1,2,3}, 
Chloé-Agathe Azencott\textsuperscript{3,1,2},
GENESIS study group\textsuperscript{\textpilcrow}
\\
\bigskip
\textbf{1} Institut Curie, PSL Research University, F-75005 Paris, France;\\
\textbf{2} INSERM, U900, F-75005 Paris, France;\\
\textbf{3} MINES ParisTech, PSL Research University, CBIO-Centre for Computational Biology, F-75006 Paris, France;\\
\textbf{4} RIKEN Center for Advanced Intelligence Project (AIP), Tokyo, Japan; \\
\textbf{5} Service de Génétique, Institut Curie, F-75005 Paris, France;\\
\textbf{6} INSERM, U830, F-75005 Paris, France;\\
\textbf{7} Université Paris Descartes.\\
\bigskip

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
% Network methods provide a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses.
% They discover susceptibility genes by jointly consider the statistical association between genetic variation and a phenotype, measured in a genome-wide association study (GWAS), and the biological context of each gene, represented as a network.
Network approaches to disease use biological networks, which model functional relationships between the molecules in a cell, to generate hypotheses about the genetics of complex diseases.
Among them, several approaches jointly consider gene scores, representing the association between each gene and the disease, and the biological context of each gene, modeled by a network.
%These approaches can select low-scoring genes if they connect multiple high-scoring genes.
Here, we obtain gene scores from genome-wide association studies (GWAS) and study six network approaches on 
% Hence, a network method could select a gene with a high P-value of association if it connects multiple low P-value genes in the network. In this work, we studied six network methods which identify subnetworks with high scores of GWAS association with a phenotype. This allows them to give more compelling results than standard SNP- and gene-level GWAS analyses, recovering causal subnetworks tightly related to cancer susceptibility.
% We apply them to
GENESIS, a GWAS on French women with non-BRCA familial breast cancer% and no pathogenic variants in \emph{BRCA1} and \emph{BRCA2}
. We provide a critical comparison of these six methods, discussing the impact of their mathematical formulation and of their parameters. % on % . Additionally, we performed an in-depth benchmarking with respect to
% the size and predictive power of their solutions, as well as on their stability and runtimes.
% Importantly,
Using a biological network yields more compelling results than standard % SNP- and gene-level
GWAS analyses. Indeed, %recovering causal subnetworks tightly related to cancer susceptibility.
we find significant overlaps between our solutions and the genes identified in the largest GWAS on breast cancer susceptibility. % Since most of the methods produce reasonable solutions, we
We further
propose to combine these solutions into a consensus network, %containing the genes selected by at least two methods. This aggregation
which
brings further insights. The consensus network contains \emph{COPS5}, a gene related to multiple hallmarks of cancer, and 14 of its neighbors. The main drawback of network methods % with respect to conventional $\chi^2$-based GWAS
is that they are not robust to small perturbations in their inputs. %To compensate for this instability, we
We therefore propose a stable consensus solution, formed by the most consistently selected genes in multiple subsamples of the data. In GENESIS, it is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than most genes. The network is organized around \emph{CUL3}, which is involved in the regulation of several genes linked to cancer progression. In conclusion, we showed how network methods help overcome the lack of statistical power of GWAS and improve their interpretation. Project-agnostic implementations of all methods are available at \url{https://github.com/hclimente/gwas-tools}. % to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an appealing explanation: only a limited number of \emph{core} genes are directly involved in the disease; but gene functions are deeply interrelated, and so many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we apply and compare six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expressions are correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,cai_broker_2010}: they tend to occupy central positions in the network (although not the most central ones); disease genes for the same pathology tend to be connected in modules; and often they are bottlenecks that interconect modules. Therefore, studying the neighborhood of disease-associated genes is effective at identifying new ones that are involved in the same biological functions \cite{huang_systematic_2018}.

@@

In this work, we studied the relevance of network-based approaches to genetics by applying these six network methods to GWAS data. They use different interpretations of the omnigenic model, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we used these network methods to identify additional breast cancer susceptibility genes. Lastly, we carried out a comparison of the solutions obtained by the different methods, and aggregated them to obtain consensus solutions of predisposition to familial breast cancer that addressed their shortcomings.

@@

We conducted association analyses in the GENESIS dataset (Section~\ref{methods:data}) at both SNP and gene levels (Section~\ref{methods:conventional}). Two genomic regions had a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (\nameref{sfig:snp_gene_manhattan}A). The former overlaps with the gene \emph{FGFR2}; the latter with \emph{CASC16} and the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1} and \emph{BRCA2} carrier studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. In GENESIS only \emph{FGFR2} was significantly associated with breast cancer at the gene-level (\nameref{sfig:snp_gene_manhattan}B).

@@

We applied six network methods to the GENESIS dataset (Section~\ref{methods:methods}). As none of the networks examined by LEAN was significant (Benjamini-Hochberg [BH] correction adjusted P\=/value $<$ 0.05), we obtained five solutions (Fig~\ref{fig:solution_overview}): one for each of the remaining four gene-based methods, and one for SConES GI (which works at the SNP level).

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant solution (BH adjusted P\=/value~$<$~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)}~Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors (Sections \ref{methods:algorithm_comparison} and \ref{methods:snp2gene}). \textbf{(B)}~Distribution of VEGAS2 P-values of the genes in the PPIN not selected by any network method (12\,213), and of those selected by 1 (575), 2 (73), or 3 (20) methods. \textbf{(C)}~Solution networks produced by the different methods. \textbf{(D)}~Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes).}
  \label{fig:solution_overview}
\end{figure}

@@

The topologies of the five solutions differ as well (Fig~\ref{fig:solution_overview}C), as measured by the median centrality  and the number of connected components (Table~\ref{tab:gene_solutions}). Three methods yield more than one connected component: SConES, as described above, SigMod, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have fewer than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM} (Section \ref{methods:bcac}).

\begin{table}[!ht]
  \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
  \centering
  \caption{
  {\bf Summary statistics on the solutions of multiple network methods on the PPIN. The first row contains the summary statistics on the whole PPIN.}}
  \begin{tabular}{lrrrrrr}
  {\bf Network } & {\bf \# genes } & {\bf \# edges } & {\bf \# components } & {\bf \mean{Betweenness} } & {\bf \median{P}\textsubscript{gene} } & {\bf \# genes in consensus } \\
  \thickhline
  HINT HT           & 13\,619 & 142\,541  & 15  & 16\,706   & 0.46  & 93/93 \\
  \hline
  dmGWAS            & 194     & 450       & 1   & 49\,115   & 0.19  & 55/93 \\
  heinz             & 4       & 3         & 1   & 113\,633  & 0.001 & 4/93  \\
  HotNet2           & 440     & 374       & 130 & 7\,739    & 0.048 & 63/93 \\
  LEAN              & 0       & 0         & 0   & -         & -     & 0/93  \\
  SConES GI         & 0 (1)   & 0         & 0   & -         & -     & 0/93  \\
  SigMod            & 142     & 249       & 11  & 92\,603   & 0.008 & 84/93 \\
  \hline
  Consensus         & 93      & 186       & 21  & 50\,737   & 0.006 & 93/93 \\
  Stable consensus  & 68      & 49        & 32  & 94\,854   & 0.005 & 43/93 \\
  \end{tabular}
  \begin{flushleft} \textbf{\# genes}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. {\bf \# components }: number of connected components. \textbf{\mean{Betweenness}}: median betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median VEGAS2 P\=/value of the selected genes. \textbf{\# genes in consensus}: number of genes in common between the method's solution, and the 93 genes in the consensus solution.
  \end{flushleft}
  \label{tab:gene_solutions}
  \end{adjustwidth}
\end{table}

Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences between the solutions of the different methods. It linked different parts of SigMod's solution to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~$<$~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) was also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). However, it additionally included submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~$<$~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) were involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~$<$~0.04).

Despite their differences, there were additional common themes. All obtained solutions had lower association P\=/values than the whole PPIN (median VEGAS2 P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}D). This illustrates the trade-off between controlling for type I error and biological relevance. However, there were nuances between solutions in this regard: heinz strongly favored genes with lower P\=/values, while dmGWAS was less conservative (median VEGAS2 P\=/values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes.

Additionally, the solutions presented other desirable properties. First, four of them were enriched in known breast cancer susceptibility genes (dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value$<$0.03). Second, the genes in three solutions displayed on average a significantly higher betweenness centrality than the rest of the genes (dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value $<$ 1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non-associated SNPs.

@@

Despite their shared properties, the differences between the solutions of the different methods suggested that each of them captured different aspects of cancer susceptibility. Indeed, out of the 668 genes that were selected by at least one method, only 93 were selected by at least two, 20 by three, and none by four or more. Encouragingly, the more methods selected a gene, the higher its association score to the phenotype (Fig \ref{fig:solution_overview}B), a relationship that plateaus at 2. Hence, to leverage on their strengths and compensate their respective weaknesses, we built a consensus solution that captured the genes shared among at least two solutions (Section~\ref{methods:methods}). This solution (Fig~\ref{fig:consensus}) contained 93 genes and exhibited the aforementioned properties of the individual solutions: enrichment in breast cancer susceptibility genes and higher betweenness centrality than the rest of the genes.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.9\linewidth]{./figures/figure_3.pdf}
  \caption{ {\bf Consensus solution on GENESIS (Section~\ref{methods:methods}) }. \textbf{(A)} ~Manhattan plot of genes; in black, the ones in the consensus solution. The Bonferroni threshold is indicated by a red line (1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(B)} Consensus network. Each gene is represented by a pie chart, which shows the methods that selected it. We enlarged the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset (Section~\ref{methods:bcac}). \textbf{(C)} The nodes are in the same disposition as in panel B, but every gene name is indicated. We colored in pink the names of genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset.}
  \label{fig:consensus}
\end{figure}

A pathway enrichment analysis of the genes in the consensus solution also showed similar pathways as the individual solutions (\nameref{stab:consensus_pwy}). We found two involved mechanisms: \emph{mitochondrial translation} and \emph{attenuation phase}. The former is supported by genes like \emph{MRPS30} (VEGAS2 P\=/value~=~0.001), which encode a mitochondrial ribosomal protein and was also linked to breast cancer susceptibility \cite{quigley_5p12_2014}. Interestingly, increased mitochondrial translation has been found in cancer cells \cite{Yu2016Repositioning}, and its inhibition proposed as a therapeutic target. With regards to the attenuation phase of heat shock response, it involved three Hsp70 chaperones: HSPA1A, HSPA1B, and HSPA1L. The genes encoding these proteins are all near each other at 6p21, in the region known as HLA. In fact, out of the 22 SNPs that map to any of these three genes, 9 mapped to all of them, and 4 to two, making it hard to disentangle their effects. \emph{HSPA1A} was the most strongly associated gene (VEGAS2 P\=/value~=~8.37~\texttimes{}~10\textsuperscript{-4}).
